---
title: "Drivers of rents"
author: "Nam Ngo - HUD"
header-includes:
  - \usepackage[default]{sourcesanspro}
mainfont: SourceSansPro
date: "`r format(Sys.time(), '%B %Y')`"
spacing: double
output:
  html_document:
    number_sections: yes
  pdf_document:
    toc: false
    number_sections: true
---

Some preliminary results for the drivers of rent projects
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(dplyr)
library(DBI)
library(tidyverse)
library(lubridate)
library(devtools)
library(readxl)
library(zoo)
library(psych) # for geometric.mean function
library(stats)
library(dbplyr)
library(questionr)
library(ggpubr)
library(lmtest)
library(cowplot)
library(ggrepel)
library(openxlsx)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
library(seasonal)
library(forecast)
library(corrplot)
library(margins)

col <- c("#003E52", "#FFC04A",  "#62B6F3", "#A4343E", "#00826E","#DEDEDE")

source("utility_functions.R")

# options(digits = 6)
options(scipen=999)

theme1 = function(base_size=12) {
  theme_minimal() +
    theme(text = element_text(size = base_size, color="grey30"),
          plot.title = element_text(size=20),
          plot.subtitle = element_text(size=10),
          plot.caption = element_text(size=10),
          legend.text = element_text(size=12),
          axis.text.y = element_text(size=10),
          axis.text.x = element_text(size=12),
          axis.title = element_text(size=12),
          strip.text = element_text (color="grey30", size=10, margin = margin(0,.5,0,.5, "cm")),
          panel.spacing = unit(1, "lines"))
}


```

```{r import-data, cache=TRUE}
combined_nz <- readRDS("data/combined_nz.RDS") 


# Add real rents measure
combined_nz <- combined_nz %>%
  arrange(desc(date)) %>%
  fill(cpi) %>%
  fill(cpi_exRent) %>%
  arrange(date) %>%
  mutate(rpi_real = rpi/cpi*1000,
         earnings_real = earnings/cpi*1000)

# Calculate changes (annual/quarterly)
t = 12

combined_nz <- combined_nz %>%
  arrange(date) %>%
  mutate(rpi.log = log(rpi)) %>%
  mutate(rpi.change = (rpi/lag(rpi,t)-1)*100) %>%
  mutate(rpi_real.change = (rpi_real/lag(rpi_real,t)-1)*100) %>%
  mutate(income.change = (income/lag(income,t)-1)*100) %>%
  mutate(earnings.change = (earnings/lag(earnings,t)-1)*100) %>%
  mutate(earnings_real.change = (earnings_real/lag(earnings_real,t)-1)*100) %>%
  mutate(earnings_sa.change = (earnings_sa/lag(earnings_sa,t)-1)*100) %>%
  mutate(dwellings.change = (dwellings/lag(dwellings,t)-1)*100) %>%
  mutate(pop.change = (pop/lag(pop,t)-1)*100) %>%
  mutate(hpi.change = (hpi/lag(hpi,t)-1)*100) %>%
  mutate(cpi.change = (cpi_exRent/lag(cpi_exRent,t)-1)*100) %>%
  mutate(ppd.demeaned = ppd - mean(ppd,na.rm = T)) %>%
  mutate(earnings.change.demeaned = earnings.change - mean(earnings.change,na.rm = T)) %>%
  mutate(income.change.demeaned = income.change - mean(income.change,na.rm = T)) %>%
  mutate(vacancy.demeaned = vacancy - mean(vacancy,na.rm = T)) %>%
  mutate(vacant_time.demeaned = vacant_time - mean(vacant_time,na.rm = T)) %>%
  mutate(mrate_lagged = mrate + lag(mrate,1) + lag(mrate,2) + lag(mrate,3)) 


combined_nz_quarterly <- combined_nz %>%
  filter(month(date) %in% c(3,6,9,12)) %>%
  #filter(year(date)>=2007) %>%
  mutate(quarter = quarter(date)) 

```

# Preliminary analysis

## Time series 

```{r ,fig.height=7, fig.width=12, warning = FALSE, message = FALSE, echo = FALSE}
# Seasonality adjustment - ARIMA-X11  
# Convert to ts object
rpi_ts <- ts(combined_nz_quarterly$rpi,freq = 4 , as.yearqtr(combined_nz_quarterly$date[1]))

# rpi_ts_sa <- seas(rpi_ts, x11 ="")

# autoplot(rpi_ts_sa)

# Decompose into trend + seasonal + residuals
autoplot(rpi_ts, main = "Rental Price Index") +
  theme1() +
  labs(y = "",
       x = "")

autoplot(diff(log(rpi_ts)), main = "Rental Price Index - quarterly change") +
  theme1() +
  labs(y = "",
       x = "")

```

## Seasonality 

There's some seasonal pattern in the Rental Price Index series, i.e high in Q1 and low in Q3, however the magnitude is not significant. A lot of the underlying seasonality has already been taken out by the fixed effects method. 

Note: 

* If quarterly inflation is used as the explanatory variable, then include quarter dummies to control for the residual seasonality pattern 
* If annual inflation is used then quarterly dummies are optional

```{r ,fig.height=7, fig.width=12, warning = FALSE, message = FALSE, echo = FALSE}

# Seasonally adjustment - X11 ARIMA
rpi_ts_sa <- seas(rpi_ts, x11 ="")

# Convert back to dataframe and bind back the raw number
rpi_sa_df <- data.frame(rpi_ts_sa$data,  date = date_to_quarter(as.Date(time(trend(rpi_ts_sa))))) %>%
  left_join(combined_nz_quarterly %>%
              select(rpi, date)) %>%
  mutate(label = month.abb[month(date)]) %>%
  mutate(label = ifelse(label %in% c("Sep","Mar"), label, NA)) %>%
  mutate(label_shocks = ifelse(date == "2020-03-31" | date == "2021-06-30", "Nationwide lockdown",
                               ifelse(date == "2007-06-30", "GFC", NA)))


# Plot seasonal pattern - to decide if seasonality adjustment is needed
ggplot(data=rpi_sa_df) +
  geom_line(aes(x = date, y = seasonal , group = 1, col = "Seasonal pattern - RPI"), size = 1) +
  labs(title = "", x="", y="", col = "") +
  theme_minimal(base_size = 18) +
  theme(panel.grid.major.x = element_blank(), panel.grid.major.y = element_line(size=0.1, color="gray92")) +
  theme(panel.grid.minor = element_blank()) +
  theme(axis.line.y = element_line(color="gray92", size = 1)) +
  theme(legend.position = "top", legend.text = element_text(size = 12)) +
  scale_x_date(date_break = "1 year", date_labels = "%y", expand = c(0, 0)) +
  theme(axis.text.x = element_text(angle = 90, size = 10),
        axis.title = element_text(size=18)) +
  scale_colour_manual(values=c("#FFBE1C"), name="") +
  geom_label_repel(aes(x = date, y = seasonal, label = label),
                   nudge_x = 80,
                   na.rm = TRUE) 

p1 <- ggplot(data=rpi_sa_df) +
  geom_line(aes(x = date, y = rpi, group = 1, alpha = "Actual"), col = col[1], size = 1) +
  geom_line(aes(x = date, y = seasonaladj, group = 2, alpha = "Seasonally adjusted"), col = col[1], size = 1.5) +
  labs(title = "Rental Price Index (seasonally adjusted)", x="", y="", col = "", fill = "") +
  theme1() +
  theme(panel.grid.major.x = element_blank(), panel.grid.major.y = element_blank()) +
  theme(panel.grid.minor = element_blank()) +
  theme(axis.line.y = element_line(color="grey92", size = 1)) +
  # Key shocks
  # Covid
  geom_text_repel(aes(x = date, y = seasonaladj, label = label_shocks),
                  nudge_x = 0.1,
                  nudge_y = -0.1,
                  na.rm = TRUE,
                  size = 4,
                  col = col[2],
                  #arrow = arrow(length = unit(0.01, "npc")),
                  box.padding = 5) +
  theme(legend.position = c(0.2, 0.9), legend.text = element_text(size = 12), legend.key.width = unit(1, 'cm')) +
  expand_limits(y=0) +
  theme(axis.text.x = element_text(angle = 45, size = 12),
        axis.title = element_text(size=18)) +
  scale_alpha_manual(values = c(0.3, 1), name = "") +
  scale_x_date(date_break = "2 year", date_labels = "%y") +
  ylim(0.5,2)

p1

```

## Lag selection - Rent growth

The autocorrelation function and partial autocorrelation function suggest that lag rent growth of at least order 1 should be included as explanatory variables

### Quarterly change

```{r ,fig.height=7, fig.width=12, warning = FALSE, message = FALSE, echo = FALSE}

t = 3

combined_nz1 <- combined_nz %>%
  arrange(date) %>%
  mutate(rpi.change = (rpi/lag(rpi,t)-1)*100)

rpi_ts <- ts(combined_nz1$rpi.change,freq = 4 , as.yearqtr(combined_nz1$date[1])) 

Acf(rpi_ts, lag.max = 6)

Pacf(rpi_ts, lag.max = 6, plot = TRUE, na.action = na.contiguous)


```

### Annual change

```{r ,fig.height=7, fig.width=12, warning = FALSE, message = FALSE, echo = FALSE}

t = 12

combined_nz1 <- combined_nz %>%
  arrange(date) %>%
  mutate(rpi.change = (rpi/lag(rpi,t)-1))

rpi_ts <- ts(combined_nz1$rpi.change,freq = 4 , as.yearqtr(combined_nz1$date[1])) 

Acf(rpi_ts, lag.max = 6)

Pacf(rpi_ts, lag.max = 6, plot = TRUE, na.action = na.contiguous)


```

### Correlation matrix


```{r ,fig.height=7, fig.width=12, warning = FALSE, message = FALSE, echo = FALSE}

combined_nz <- readRDS("data/combined_nz.RDS") 

# Add real rents measure
combined_nz <- combined_nz %>%
  arrange(desc(date)) %>%
  fill(cpi) %>%
  fill(cpi_exRent) %>%
  arrange(date) %>%
  mutate(rpi_real = rpi/cpi*1000,
         earnings_real = earnings/cpi*1000)

# Calculate changes (annual/quarterly)
t = 12

combined_nz <- combined_nz %>%
  arrange(date) %>%
  mutate(rpi.log = log(rpi)) %>%
  mutate(rpi.change = (rpi/lag(rpi,t)-1)*100) %>%
  mutate(rpi_real.change = (rpi_real/lag(rpi_real,t)-1)*100) %>%
  mutate(income.change = (income/lag(income,t)-1)*100) %>%
  mutate(earnings.change = (earnings/lag(earnings,t)-1)*100) %>%
  mutate(earnings_real.change = (earnings_real/lag(earnings_real,t)-1)*100) %>%
  mutate(earnings_sa.change = (earnings_sa/lag(earnings_sa,t)-1)*100) %>%
  mutate(dwellings.change = (dwellings/lag(dwellings,t)-1)*100) %>%
  mutate(pop.change = (pop/lag(pop,t)-1)*100) %>%
  mutate(hpi.change = (hpi/lag(hpi,t)-1)*100) %>%
  mutate(cpi.change = (cpi_exRent/lag(cpi_exRent,t)-1)*100) %>%
  mutate(ppd.demeaned = ppd - mean(ppd,na.rm = T)) %>%
  mutate(earnings.change.demeaned = earnings.change - mean(earnings.change,na.rm = T)) %>%
  mutate(income.change.demeaned = income.change - mean(income.change,na.rm = T)) %>%
  mutate(vacancy.demeaned = vacancy - mean(vacancy,na.rm = T)) %>%
  mutate(vacant_time.demeaned = vacant_time - mean(vacant_time,na.rm = T)) %>%
  mutate(mrate_lagged = mrate + lag(mrate,1) + lag(mrate,2) + lag(mrate,3)) 

combined_nz_quarterly <- combined_nz %>%
  filter(month(date) %in% c(3,6,9,12)) %>%
  #filter(year(date)>=2007) %>%
  mutate(quarter = quarter(date)) 


cor_matrix <- cor(combined_nz_quarterly %>%
      filter(year(date)>=2007) %>%
      dplyr::select(rpi.change,
             income.change,
             earnings.change,
             ppd.demeaned,
             unemp,
             mrate))

# round(cor_matrix, 2)

corrplot(cor_matrix, method="number")



```


## Results

### Annual rent inflation, quarterly frequency vs wage growth (average weekly earnings)

```{r Prepare data}

combined_nz <- readRDS("data/combined_nz.RDS") 

# Add real rents measure
combined_nz <- combined_nz %>%
  arrange(desc(date)) %>%
  fill(cpi) %>%
  fill(cpi_exRent) %>%
  arrange(date) %>%
  mutate(rpi_real = rpi/cpi*1000,
         earnings_real = earnings/cpi*1000)

# Calculate changes (annual/quarterly)
t = 12

combined_nz <- combined_nz %>%
  arrange(date) %>%
  mutate(rpi.change = (rpi/lag(rpi,t)-1)*100) %>%
  mutate(rpi_real.change = (rpi_real/lag(rpi_real,t)-1)*100) %>%
  mutate(income.change = (income/lag(income,t)-1)*100) %>%
  mutate(earnings.change = (earnings/lag(earnings,t)-1)*100) %>%
  mutate(earnings_real.change = (earnings_real/lag(earnings_real,t)-1)*100) %>%
  mutate(earnings_sa.change = (earnings_sa/lag(earnings_sa,t)-1)*100) %>%
  mutate(dwellings.change = (dwellings/lag(dwellings,t)-1)*100) %>%
  mutate(pop.change = (pop/lag(pop,t)-1)*100) %>%
  mutate(hpi.change = (hpi/lag(hpi,t)-1)*100) %>%
  mutate(cpi.change = (cpi_exRent/lag(cpi_exRent,t)-1)*100) %>%
  mutate(ppd.demeaned = ppd - mean(ppd,na.rm = T)) %>%
  mutate(earnings.change.demeaned = earnings.change - mean(earnings.change,na.rm = T)) %>%
  mutate(income.change.demeaned = income.change - mean(income.change,na.rm = T)) %>%
  mutate(vacancy.change = (vacancy/lag(vacancy,t)-1)*100) %>%

  mutate(vacancy.demeaned = vacancy - mean(vacancy,na.rm = T)) %>%
  mutate(vacant_time.demeaned = vacant_time - mean(vacant_time,na.rm = T)) %>%
  mutate(vacant_time.change = vacant_time - lag(vacant_time,t)) %>%

  mutate(mrate_ave = (lag(mrate,0) + lag(mrate,1) + lag(mrate,2))/3) %>%
  #mutate(mrate.change = (mrate/lag(mrate,t)-1)*100) %>%
  mutate(mrate.change = mrate - lag(mrate,t)) %>%

  mutate(covid_dummy = ifelse(year(date) %in% c(2020,2021), 1, 0),
         gfc_dummy = ifelse(year(date) %in% c(2007,2008),1,0)) 

combined_nz_quarterly <- combined_nz %>%
  filter(month(date) %in% c(3,6,9,12)) %>%
  #filter(year(date)>=2007) %>%
  mutate(quarter = quarter(date)) 


# combined_nz_quarterly <- combined_nz %>%
#   filter(month(date) %in% c(3,6,9,12)) %>%
#   #filter(year(date)>=2007) %>%
#   mutate(quarter = quarter(date)) %>%
#   mutate(lag.rpi = lag(rpi.change,1),
#          lag.earnings = lag(earnings.change,1),
#          lag.ppd = lag(ppd.demeaned,1),
#          lag.mrate = lag(mrate.change,1),
#          lag.unemp = lag(unemp,1))

```


* Measure of relative supply and demand: People per dwelling (demeanded) 
* Q1-2006 to Q4-2021


```{r Baseline model 1}}


model <- lm(rpi.change ~ 
              lag(rpi.change,1) +
              lag(earnings.change,1) + 
              lag(ppd.demeaned,1) +
              lag(ppd.demeaned,1)*lag(earnings.change,1) +

              lag(mrate,1) + 


              
              #lag(pop.change,1) + 
              #lag(dwellings.change,1) + 
              
              #lag(cpi.change,1) +
              lag(unemp,1) +
              factor(quarter) 

              #factor(year(date))
            ,
            data = combined_nz_quarterly) 


# model <- lm(rpi.change ~ 
#               lag.rpi +
#              lag.earnings + 
#               lag.ppd +
#               lag.ppd*lag.earnings +
# 
#               lag.mrate +
#               
#               #lag(pop.change,1) + 
#               #lag(dwellings.change,1) + 
#               
#               #cpi.change +
#               lag.unemp +
#               factor(quarter)
#               #factor(year(date))
#             ,
#             data = combined_nz_quarterly) 


labels <- c("Rent inflation (lagged)",
           "Wage growth (lagged)",
           "People per dwelling (demeaned, lagged)",
           "Floating mortgage rate (lagged)",
           "Unemployment rate (lagged)",
           "Wage growth x People per dwelling")

tab_model(summary(model),   
          show.intercept = F,
          digits = 3,
          dv.labels = "Rent inflation (annual)",
          pred.labels = labels,
          drop = "quarter",
          p.style = "numeric_stars",
          p.threshold = c(0.1, 0.05, 0.01))  


```

Replacing mortgage rate with change in mortgage rate suggests that the speed of rising/decreasing mortgage rate is more important for explaining rent inflation (through declining/rising homeownership rate channel)

```{r Baseline model 1b}}


model <- lm(rpi.change ~ 
              lag(rpi.change,1) +
              lag(earnings.change,1) + 
              lag(ppd.demeaned,1) +
              lag(ppd.demeaned,1)*lag(earnings.change,1) +

              lag(mrate.change,1) + 


              
              #lag(pop.change,1) + 
              #lag(dwellings.change,1) + 
              
              #lag(cpi.change,1) +
              lag(unemp,1) +
              factor(quarter) 

              #factor(year(date))
            ,
            data = combined_nz_quarterly) 


# model <- lm(rpi.change ~ 
#               lag.rpi +
#              lag.earnings + 
#               lag.ppd +
#               lag.ppd*lag.earnings +
# 
#               lag.mrate +
#               
#               #lag(pop.change,1) + 
#               #lag(dwellings.change,1) + 
#               
#               #cpi.change +
#               lag.unemp +
#               factor(quarter)
#               #factor(year(date))
#             ,
#             data = combined_nz_quarterly) 


labels <- c("Rent inflation (lagged)",
           "Wage growth (lagged)",
           "People per dwelling (demeaned, lagged)",
           "Change in floating mortgage rate (lagged)",
           "Unemployment rate (lagged)",
           "Wage growth x People per dwelling")

tab_model(summary(model),   
          show.intercept = F,
          digits = 3,
          dv.labels = "Rent inflation (annual)",
          pred.labels = labels,
          drop = "quarter",
          p.style = "numeric_stars",
          p.threshold = c(0.1, 0.05, 0.01))  


```

### Alternative measure of relative supply and demand

* Measure of relative supply and demand: Vacancy rate (New bonds/Active bonds) 
* Q1-2000 to Q4-2021


```{r Baseline model 2}}


model <- lm(rpi.change ~ 
              lag(rpi.change,1) +
              lag(earnings.change,1) + 
              lag(vacancy,1) +
              lag(vacancy,1)*lag(earnings.change,1) +

              lag(mrate,1) +
              
              #lag(pop.change,1) + 
              #lag(dwellings.change,1) + 
              
              #cpi.change +
              lag(unemp,1) +
              factor(quarter) 
              #factor(year(date))
            ,
            data = combined_nz_quarterly) 



labels <- c("Rent inflation (lagged)",
           "Wage growth (lagged)",
           "Vacancy rate (lagged)",
           "Change in floating mortgage rate (lagged)",
           "Unemployment rate (lagged)",
           "Wage growth x Vacancy rate")

tab_model(summary(model),   
          show.intercept = F,
          digits = 3,
          dv.labels = "Rent inflation (annual)",
          pred.labels = labels,
          drop = "quarter",
          p.style = "numeric_stars",
          p.threshold = c(0.1, 0.05, 0.01)) 

```


* Measure of relative supply and demand: Change in median vacant time (Time between tenancies) 
* Q1-2000 to Q4-2021


```{r Baseline model 3}}


model <- lm(rpi.change ~ 
              lag(rpi.change,1) +
              lag(earnings.change,1) + 
              lag(vacant_time.change,1) +
              lag(vacant_time.change,1)*lag(earnings.change,1) +

              lag(mrate,1) +
              
              #lag(pop.change,1) + 
              #lag(dwellings.change,1) + 
              
              #cpi.change +
              lag(unemp,1) +
              factor(quarter) 
              #factor(year(date))
            ,
            data = combined_nz_quarterly) 


# model <- lm(rpi.change ~ 
#               lag.rpi +
#              lag.earnings + 
#               lag.ppd +
#               lag.ppd*lag.earnings +
# 
#               lag.mrate +
#               
#               #lag(pop.change,1) + 
#               #lag(dwellings.change,1) + 
#               
#               #cpi.change +
#               lag.unemp +
#               factor(quarter)
#               #factor(year(date))
#             ,
#             data = combined_nz_quarterly) 


labels <- c("Rent inflation (lagged)",
           "Wage growth (lagged)",
           "Change in median vacant time (lagged)",
           "Change in floating mortgage rate (lagged)",
           "Unemployment rate (lagged)",
           "Wage growth x Change in median vacant time")

tab_model(summary(model),   
          show.intercept = F,
          digits = 3,
          dv.labels = "Rent inflation (annual)",
          pred.labels = labels,
          drop = "quarter",
          p.style = "numeric_stars",
          p.threshold = c(0.1, 0.05, 0.01)) 

```


Household income growth (to be added)

```{r Model 2}

# model <- lm(rpi.change ~ 
#               lag(income.change,1) +
#               lag(earnings.change,1) + 
#               ppd.demeaned +
#               ppd.demeaned*lag(income.change,1) +
# 
#               mrate +
#               
#               #lag(pop.change,1) + 
#               #lag(dwellings.change,1) + 
#               
#               #cpi.change +
#               lag(unemp,1) +
#               factor(quarter) 
#               #factor(year(date))
#             ,
#             data = combined_nz_quarterly) 
# 
# 
# labels <- c("Rent inflation (lagged)",
#            "Income growth (lagged)",
#            "People per dwelling (demeaned)",
#            "Mortgage rate (floating)",
#            "Unemployment rate (lagged)",
#            "Wage growth x People per dwelling")
# 
# tab_model(summary(model),   
#           show.intercept = F,
#           digits = 3,
#           dv.labels = "Rent inflation (annual)",
#           pred.labels = labels,
#           drop = "quarter",
#           p.style = "numeric_stars",
#           p.threshold = c(0.05, 0.01, 0.001))  

```
